Temporal ordering and registration of imaging data in cross--sectional studies of development

### 2014-04-04 image_analysis_paper/pnas_draft.tex
added henry's comments

\author{Carmeline J. Dsilva\affil{1}{Department of Chemical and Biological Engineering, Princeton University, Princeton, NJ, 08544}, 
Bomyi Lim\affil{1}{}, Stanislav Y. Shvartsman\affil{1}{} \affil{2}{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, NJ 08544}, \and Ioannis G. Kevrekidis\affil{1}{} \affil{3}{Program in Applied and Computational Mathematics, Princeton University, Princeton, NJ 08544}}

@@

\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal and cross-sectional. 
%
In longitudinal studies, developmental progress is monitored over time within the same embryo. 
%
In a cross-sectional study, one embryo contributes only one snapshot of a chemical or morphological process along developmental trajectory which must be reconstructed from multiple snapshots of different embryos. 
%
Both of these experimental designs have their advantages and limitations and are extensively used by developmental biologists. 
%
Here we focus on cross-sectional studies, which have a time-honored history and still present the only option for most organisms. 
%
In a typical cross-sectional study, a group of developing embryos is fixed using a procedure that arrests their development and stained with chemicals that visualize a handful of cellular processes. 
%
Fixed embryos are then imaged using any given number of microscopic techniques. 
%
Recent advances in large scale physical manipulation of embryos and their imaging produce rapidly increasing volumes of data from such experiments, in which every embryo is observed at a different orientation and at a different time point.
%
Importantly, the ``age'' of any given embryo arrested in its development is not known to high accuracy. 
%
In general, it is known that a collection of embryos belongs to a certain time window.
%
In order to mine such datasets for underlying developmental dynamics, snapshots of different embryos must be spatially aligned and ordered in time. 
%
We present a general framework that greatly accelerates both of these tasks.

Temporal ordering and registration of images is straightforward when the number of images is small and differences between them are apparent. 
%
As an example, Figure~\ref{fig:fish} shows a schematic of fish development, combining the processes of growth and patterning.  
%
In this case, temporal ordering can be accomplished by arranging the fish by size, which correlates with the progress of development, and image registration is based on obvious morphological landmarks, such as positions of head and fins. 
%
Dealing with real data is far more nontrivial. 
%
To illustrate most of the issues associated with temporal ordering and registration of images, we use a data set from an imaging study of pattern formation in the early fruit fly embryo, one of the leading systems for studies of developmental dynamics. 
%
This dataset comprises $81$ fluorescence microscopy images of different embryos fixed during the $3^{rd}$ hour of their development.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig1}
\caption{Schematic illustration of reconstructing dynamics from cross-sectional data. (a) Fish, each in a different orientation and a different stage of development. (b) Fish, now registered and temporally ordered. For this example, the registration and ordering is easy to do by hand because the data set is small and the developmental changes are easy to recognize.} 
%\label{fig:fish}
\customlabel{fig:fish}{1}
\customlabel{subfig:fish_unordered}{\ref{fig:fish}a}
\customlabel{subfig:fish_ordered}{\ref{fig:fish}b}
\end{figure}

@@

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig2}
\caption{(a) Fluorescent images of {\em Drosophila} embryos. Each embryo has been stained from Dorsal (green) and dpERK (red). (b) Selected image from data set. (c) Profile of Dorsal (green) and dpERK (red) around the circumfrence of the embryo in (b).}
%\label{fig:images_unordered}
\customlabel{fig:images_unordered}{2}
\customlabel{subfig:images_unordered}{\ref{fig:images_unordered}a}
\customlabel{subfig:one_image}{\ref{fig:images_unordered}b}
\customlabel{subfig:image_profiles}{\ref{fig:images_unordered}c}
\end{figure}

@@

We will first demonstrate our techniques for registration and temporal ordering using a synthetic data set with relatively simple dynamics.
%
We construct concentration profiles defined on a ring, and we assume that each ring is in an arbitrary orientation; an example is shown in Figure~\ref{subfig:1d_example}.
%
Each concentration profile is a noisy Gaussian (shown in Figure~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
%
We discritize the profiles into 100 points, so our data will be 100-dimensional vectors. 
%
Rotation of the ring in Figure~\ref{subfig:1d_example} corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown below the ring. 
%
Figure \ref{subfig:1d_unaligned_unordered} shows all the concentration profiles in our data set. 
%
The concentration profiles have been stacked in an array, such that each row corresponds to one data point.
%
Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian growing in time) are not readily apparent.

\begin{figure}[t]
%\includegraphics{figures_0329_mac_4}
\includegraphics[width=8.4cm]{fig4}
\caption{Illustration of methodologies using one-dimensional profiles. (a) One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom). (b) Intensity corresponding to the profile in (a). (c) Many concentration profiles, each of the form described in (a). Each row in the array corresponds to one image. (d) The concentration profiles in (b), now registered using angular synchronization. (e) The concentration profiles in (c), now temporally ordered using diffusion maps. (f) The concentration profiles in (b), now simultaneously registered and temporally ordered using vector diffusion maps.}
%\label{fig:1d_demo}
\customlabel{fig:1d_demo}{3}
\customlabel{subfig:1d_example}{\ref{fig:1d_demo}a}
\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}b}
\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}c}
\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}d}
\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}e}
\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}f}
\end{figure}

@@

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig3}
\caption{Illustration of diffusion maps and angular synchronization (a) Data points (in blue) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of data points is connected by an edge, and the edge intensity is related to the distance between the points (close data points are connected by dark edges). (b) The data in (a), colored by the first (non-trivial) eigenvector from diffusion maps, $\phi_2$. The color parameterizes the one-dimensional curve. (c) Set of vectors, each in a different orientation. The angles to align one vector to another are indicated. (d) The vectors from (c), each rotated so that the set of vectors is aligned. Note that the angles are consistent with the pairwise alignments in (c). }
%\label{fig:schematics}
\customlabel{fig:schematics}{4}
\customlabel{subfig:dmaps_edges}{\ref{fig:schematics}a}
\customlabel{subfig:dmaps_color}{\ref{fig:schematics}b}
\customlabel{subfig:synch1}{\ref{fig:schematics}c}
\customlabel{subfig:synch2}{\ref{fig:schematics}d}
\end{figure}

@@

\begin{figure}[t]
%\includegraphics{figures_0329_mac_6}
\includegraphics[width=8.4cm]{fig6}
\caption{(a) Images, stained for membranes (gray) and Dorsal (green) at different levels of development. (b) Membrane length as a function of time. (c) The one-dimensional concentration profiles, extracted, registered, and ordered by hand. The profiles are extracted using the nuclei positions in the image, registered using the location of the Dorsal peak, and ordered using cellularization. (d) The one-dimensional concentration profiles corresponding to the images in Figure~\ref{fig:images_ordered}. }
%\label{fig:membrane_compare}
\customlabel{fig:membrane_compare}{6}
\customlabel{subfig:membranes}{\ref{fig:membrane_compare}a}
\customlabel{subfig:membrane_curve}{\ref{fig:membrane_compare}b}
\customlabel{subfig:1d_membranes}{\ref{fig:membrane_compare}c}
\customlabel{subfig:1d_vdm}{\ref{fig:membrane_compare}d}
\end{figure}

@@

We note that, under our assumption that $\theta_{ij} \approx \theta_i - \theta_j$, 
\begin{equation} 
H_{ij} = R(\theta_{ij}) \approx R(\theta_i - \theta_j) = R(\theta_i) R(-\theta_j) = R(\theta_i) R(\theta_j)^T,
\end{equation}
 and so
\begin{equation} \label{eq:H_low_rank}
	H \approx 
	\begin{bmatrix}
	R(\theta_1) \\
	R(\theta_2) \\
	\vdots \\
	R(\theta_m)
	\end{bmatrix}
	\begin{bmatrix}
	R(\theta_1)^T R(\theta_2)^T \dots R(\theta_m)^T
	\end{bmatrix}.
\end{equation}
%
From \eqref{eq:H_low_rank}, we can see that the top block eigenvector of $H$, which we will denote $\hat{R}$, contains estimates of $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$.
%
Let $\phi_1, \phi_2, \dots, \phi_{md}$ denote the eigenvectors of $H$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$, where $\lambda_i$ is the eigenvalue corresponding to $\phi_i$. 
%
Then,
\begin{equation}
\hat{R} = 
\begin{bmatrix}
\hat{R}_1 \\
\hat{R}_2 \\
\vdots \\
\hat{R}_m
\end{bmatrix} =
\begin{bmatrix}
| & | & & | \\
\phi_1 & \phi_2 & \dots & \phi_d \\
| & | & & | 
\end{bmatrix},
\end{equation}
and $\hat{R}_i \in \mathbb{R}^{d \times d}$ is the estimate for $R(\theta_i)$. 
%
However, $\hat{R}_i$ is not guaranteed to be a rotation matrix; the closest rotation matrix, $R_{i, est}$, is given by
\begin{equation} \label{eq:R_est}
R_{i, est} = U_i V_i^T, 
\end{equation} 
where $U_i$ and $V_i$ are the left and right singular vectors, respectively, of $\hat{R}_i$.
%
We adjust the sign of $\phi_1$ so that $det(R_{i, est}) = +1$, ensuring proper rotations.
%
We estimate $\theta_{i}$ by inverting \eqref{eq:R_theta}, and register the signals by rotating signal $i$ by $-\theta_i$.

@@

We then compute the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_{md}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{md}$ of $S$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$.
%
As in angular synchronization, the top (block) eigenvector of $S$ contains approximations of the optimal rotations, and the optimal rotations can be computed in the same way from \eqref{eq:R_est}.
%
However, the eigenvectors now also contain information about the embedding coordinates for our images.
%
In general, the embedding coordinates are given by 
\begin{equation}
\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle
\end{equation}
where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$.
%
If we assume that our data are one-dimensional, and that the rotations and the dynamics are uncoupled and therefore separable, the embedding coordinate for our data (i.e., the analog of $\phi_2$ from the diffusion maps case) is given by $\psi_{1,d+1}$.
%
If no such claims can be made, then one often runs a second step of diffusion maps, using $\psi_{k,l}$ as the coordinates, to uncover a parameterization of the manifold modulo symmetries.

@@

%As discussed above, for angular synchronization, the underlying symmetry group needs to have a real and orthogonal matrix representation. 
%
%We therefore convert $ISO(2)$, the group of two-dimensional translations and rotations, to $SO(3)$, the group of three-dimensional rotations. 
%
We then convert the two-dimensional translation and rotation to a three dimensional rotation. 
%
We first compute the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$,
\begin{equation} \label{eq:angle_relations}
\begin{aligned}
	\alpha_{ij} &= \theta_{ij} \\
	\beta_{ij} &= \frac{dx_{ij}}{n} \times \eta_{proj} \\
	\gamma_{ij} &= \frac{dy_{ij}}{n} \times \eta_{proj} \\
\end{aligned}
\end{equation}
where $\eta_{proj}$ is the angular portion of the sphere onto which we choose to project the image.
%
We take $\eta_{proj} =  \pi/8$, so the image lies on a $\pi/8 \times \pi/8$ radians portion of the unit sphere.
%
This portion of the sphere is small enough so that curvature effects are not seen in the image.
%
We write rotations around the three principal axes, $R^x(\alpha)$, $R^y(\beta)$, and $R^z(\gamma)$, in terms of the three Euler angles
%\small
\begin{equation}
\begin{aligned}
	R^x(\alpha) &= \begin{bmatrix}
	1 & 0 & 0 \\
    0 & \cos(\alpha) & -\sin(\alpha) \\
    0 & \sin(\alpha) & \cos(\alpha)
	\end{bmatrix} \\
	R^y(\beta) &= \begin{bmatrix}
	\cos(\beta) & 0 & \sin(\beta) \\
    0 & 1 & 0 \\
    -\sin(\beta) & 0 & \cos(\beta)
    \end{bmatrix} \\
	R^z(\gamma) &= \begin{bmatrix} 
	\cos(\gamma) & -\sin(\gamma) & 0 \\
    \sin(\gamma) & \cos(\gamma) & 0 \\
    0 & 0 & 1 
    \end{bmatrix}.
\end{aligned}
\end{equation}
%\normalsize
%
The total rotation, $R_{ij} \in SO(3)$, is 
\begin{equation} \label{eq:total_rot}
	R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})	 = R^z(\gamma_{ij})  R^y(\beta_{ij})  R^x(\alpha_{ij})
%R(\alpha_{ij}, \beta_{ij}, \gamma_{ij}) = 
%\begin{bmatrix}
%\cos(\beta) \cos(\gamma) & \cos(\gamma)\sin(\alpha)\sin(\beta)-\cos(\alpha) \sin(\gamma) & \cos(\alpha)\cos(\gamma)\sin(\beta)+ \sin(\alpha)\sin(\gamma) \\
%\cos(\beta) \sin(\gamma) & \cos(\alpha) \cos(\gamma)+\sin(\alpha) \sin(\beta) \sin(\gamma) & \cos(\alpha) \sin(\beta) \sin(\gamma)-\cos(\gamma) \sin(\alpha) \\
%-\sin(\beta) & \cos(\beta) \sin(\alpha) & \cos(\alpha) \cos(\beta))
%\end{bmatrix}
\end{equation}
%
%\eqref{eq:total_rot} corresponds to first rotating the image by $\theta$, then translating the image by $dx$ pixels in the $x$ direction, and finally translating the image by $dy$ pixels in the $y$ direction (we note that, because the rotation matrices operate from the left, the rightmost rotation matrix in the product in \eqref{eq:total_rot} corresponds to the first operation we perform on the image).

From the rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$,
we build the matrix $H$ in \eqref{eq:H_to_R} and compute the estimates $R_{i, est}$ in \eqref{eq:R_est}.
%
From the matrices $R_{1, est}, \dots, R_{m, est}$, we then find the corresponding translations and rotations of the images.
%
We first multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
%
We then compute the Euler angles $\alpha$, $\beta$, and $\gamma$ from the rotation matrix $R$ using the following relationships
\begin{equation}
\begin{aligned}
R_{1,1} & = \cos(\beta)\cos(\gamma) \\
R_{2,1} & = \cos(\beta)\sin(\gamma) \\
R_{3,1} & = -\sin(\beta) \\
R_{3,2} & = \sin(\alpha)\cos(\beta) \\
R_{3,3} & = \cos(\alpha)\cos(\beta) 
\end{aligned}
\end{equation}
%
We then invert \eqref{eq:angle_relations} to compute the optimal translation and rotation for each image.
%
We round the optimal translations to the nearest integer to remove the need for interpolation when translating the images.

>>>>

\author{Carmeline~J.~Dsilva\affil{1}{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA},
Bomyi~Lim\affil{1}{}, 
Thomas~J.~Levario\affil{2}{School of Chemical and Biomolecular Engineering, Georgia Institute of Technology, Atlanta, Georgia, USA},
Hang~Lu\affil{2}{},
Stanislav~Y.~Shvartsman\affil{1}{} \affil{3}{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA}, 
\and 
Ioannis~G.~Kevrekidis\affil{1}{} \affil{4}{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA}}

@@

\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal and cross-sectional. 
%
In longitudinal studies, developmental progress is monitored over time within the same embryo. 
%
In a cross-sectional study, one embryo contributes only one snapshot of a chemical or morphological process along a developmental trajectory, which must be reconstructed from multiple snapshots of different embryos. 
%
Both of these experimental designs have their advantages and limitations and are extensively used by developmental biologists. 
%
Here we focus on cross-sectional studies, which have a time-honored history and still present the only option for most organisms. 
%
In a typical cross-sectional study, a group of developing embryos is fixed using a procedure that arrests their development and stained with chemicals that visualize a handful of cellular processes. 
%
Fixed embryos are then imaged using any given number of microscopic techniques. 
%
Recent advances in large-scale physical manipulation and imaging of embryos have produced rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different orientation and time point.
%
Importantly, the ``age'' of any given embryo arrested in its development is not known to high accuracy. 
%
In general, it is known that a collection of embryos belongs to a certain time window.
%
In order to mine such datasets for underlying developmental dynamics, snapshots of different embryos must be spatially aligned and ordered in time. 
%
We present a general framework that greatly accelerates both of these tasks.

Temporal ordering and registration of images is straightforward when the number of images is small and differences between them are apparent. 
%
As an example, Figure~\ref{fig:fish} shows a schematic of fish development, combining the processes of growth and patterning.  
%
In this case, temporal ordering can be accomplished by arranging the fish by size, which correlates with the progress of development, and image registration is based on obvious morphological landmarks, such as positions of head and fins. 
%
Dealing with real data is far more nontrivial, due to the number of images and similarities between them. 
%
To illustrate most of the issues associated with temporal ordering and registration of images, we use a data set from an imaging study of pattern formation in the early fruit fly embryo, one of the leading systems for studies of developmental dynamics. 
%
This dataset comprises $49$ fluorescence microscopy images of different embryos fixed during the $3^{rd}$ hour of their development.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig1}
\caption{Schematic illustration of reconstructing dynamics from cross-sectional data. (a) Fish, each in a different orientation and a different stage of development. (b) Fish, now registered and temporally ordered by hand. For this example, the registration and ordering is easy to do by hand because the data set is small and the developmental changes are easy to recognize.} 
%\label{fig:fish}
\customlabel{fig:fish}{1}
\customlabel{subfig:fish_unordered}{\ref{fig:fish}a}
\customlabel{subfig:fish_ordered}{\ref{fig:fish}b}
\end{figure}

@@

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig2}
\caption{(a) Fluorescent images of {\em Drosophila} embryos. Each embryo has been stained for Dorsal (green) and dpERK (red). (b) Selected image from data set. (c) Profile of Dorsal (green) and dpERK (red) around the circumfrence of the embryo in (b).}
%\label{fig:images_unordered}
\customlabel{fig:images_unordered}{2}
\customlabel{subfig:images_unordered}{\ref{fig:images_unordered}a}
\customlabel{subfig:one_image}{\ref{fig:images_unordered}b}
\customlabel{subfig:image_profiles}{\ref{fig:images_unordered}c}
\end{figure}

@@

We will first demonstrate our techniques for registration and temporal ordering using a synthetic data set with relatively simple dynamics.
%
We construct concentration profiles defined on a ring, and we assume that each ring is in an arbitrary orientation; an example is shown in Figure~\ref{subfig:1d_example}.
%
Each concentration profile is a noisy Gaussian (shown in Figure~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
%
We discritize the profiles into 100 points, so our data will be 100-dimensional vectors. 
%
Rotation of the ring in Figure~\ref{subfig:1d_example} corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown below the ring. 
%
Figure \ref{subfig:1d_unaligned_unordered} shows all the concentration profiles in our data set. 
%
The concentration profiles have been stacked in an array, such that each row corresponds to one profile.
%
Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian growing in time) are not readily apparent.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig3}
\caption{Illustration of methodologies using one-dimensional profiles. (a) One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom). (b) Intensity corresponding to the profile in (a). (c) Many concentration profiles, each of the form described in (a). Each row in the array corresponds to one image. (d) The concentration profiles in (c), now registered using angular synchronization. (e) The concentration profiles in (c), now temporally ordered using diffusion maps. (f) The concentration profiles in (c), now simultaneously registered and temporally ordered using vector diffusion maps.}
%\label{fig:1d_demo}
\customlabel{fig:1d_demo}{3}
\customlabel{subfig:1d_example}{\ref{fig:1d_demo}a}
\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}b}
\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}c}
\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}d}
\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}e}
\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}f}
\end{figure}

@@

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig4}
\caption{(a) Set of vectors, each in a different orientation. The angles to align one vector to another are indicated. (b) The vectors from (a), each rotated so that the set of vectors is aligned. Note that the angles are consistent with the pairwise alignments in (a). (c) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of data points is connected by an edge, and the edge intensity is related to the distance between the points (close data points are connected by dark edges). (d) The data in (c), colored by the first (non-trivial) eigenvector from diffusion maps, $\phi_2$. The color parameterizes the one-dimensional curve.  }
%\label{fig:schematics}
\customlabel{fig:schematics}{4}
\customlabel{subfig:dmaps_edges}{\ref{fig:schematics}a}
\customlabel{subfig:dmaps_color}{\ref{fig:schematics}b}
\customlabel{subfig:synch1}{\ref{fig:schematics}c}
\customlabel{subfig:synch2}{\ref{fig:schematics}d}
\end{figure}

@@

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig6}
\caption{(a) Images, stained for membranes (gray) and Dorsal (green) at different levels of development. (b) Membrane length as a function of time. (c) The one-dimensional concentration profiles, registered by peak Dorsal fluorescence and ordered by membrane length. (d) The one-dimensional concentration profiles corresponding to the automatically registered and ordered images in Figure~\ref{fig:images_ordered}. }
%\label{fig:membrane_compare}
\customlabel{fig:membrane_compare}{6}
\customlabel{subfig:membranes}{\ref{fig:membrane_compare}a}
\customlabel{subfig:membrane_curve}{\ref{fig:membrane_compare}b}
\customlabel{subfig:1d_membranes}{\ref{fig:membrane_compare}c}
\customlabel{subfig:1d_vdm}{\ref{fig:membrane_compare}d}
\end{figure}

@@

We note that, under our assumption that $\theta_{ij} \approx \theta_i - \theta_j$, $H_{ij} \approx R(\theta_i) R(\theta_j)^T$,
%\begin{equation} 
%H_{ij} = R(\theta_{ij}) \approx R(\theta_i - \theta_j) = R(\theta_i) R(-\theta_j) = R(\theta_i) R(\theta_j)^T,
%\end{equation}
 and so
\begin{equation} \label{eq:H_low_rank}
	H \approx 
	\begin{bmatrix}
	R(\theta_1) \\
	R(\theta_2) \\
	\vdots \\
	R(\theta_m)
	\end{bmatrix}
	\begin{bmatrix}
	R(\theta_1)^T R(\theta_2)^T \dots R(\theta_m)^T
	\end{bmatrix}.
\end{equation}
%
From \eqref{eq:H_low_rank}, we can see that the top block eigenvector of $H$, which we will denote $\hat{R}$, contains estimates of $R(\theta_1), R(\theta_2), \dots, R(\theta_m)$.
%
Let $\phi_1, \phi_2, \dots, \phi_{md}$ denote the eigenvectors of $H$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$, where $\lambda_i$ is the eigenvalue corresponding to $\phi_i$. 
%
Then,
\begin{equation}
\hat{R} = 
\begin{bmatrix}
\hat{R}_1 \\
\hat{R}_2 \\
\vdots \\
\hat{R}_m
\end{bmatrix} =
\begin{bmatrix}
| & | & & | \\
\phi_1 & \phi_2 & \dots & \phi_d \\
| & | & & | 
\end{bmatrix},
\end{equation}
and $\hat{R}_i \in \mathbb{R}^{d \times d}$ is the estimate for $R(\theta_i)$. 
%
However, $\hat{R}_i$ is not guaranteed to be a rotation matrix; the closest rotation matrix, $R_{i, est}$, is given by
\begin{equation} \label{eq:R_est}
R_{i, est} = U_i V_i^T, 
\end{equation} 
where $U_i$ and $V_i$ are the left and right singular vectors, respectively, of $\hat{R}_i$.
%
We adjust the sign of $\phi_1$ so that $det(R_{i, est}) = +1$, ensuring proper rotations.
%
We estimate $\theta_{i}$ by inverting \eqref{eq:R_theta}, and register the signals by rotating signal $i$ by $-\theta_i$.

@@

We then compute the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_{md}$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_{md}$ of $S$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_{md}|$.
%
As in angular synchronization, the top (block) eigenvector of $S$ contains approximations of the optimal rotations, and the optimal rotations can be computed in the same way from \eqref{eq:R_est}.
%
However, the eigenvectors now also contain information about the embedding coordinates for our images.
%
In general, the embedding coordinates are given by 
\begin{equation}
\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle
\end{equation}
where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$.
%
If we assume that our data are one-dimensional, and that the rotations and the dynamics are uncoupled and therefore separable, the embedding coordinate for our data %(i.e., the analog of $\phi_2$ from the diffusion maps case) 
is given by $\psi_{1,d+1}$.
%
If no such claims can be made, then one often runs a second step of diffusion maps, using $\psi_{k,l}$ as the coordinates, to uncover a parameterization of the manifold modulo symmetries.

@@

%As discussed above, for angular synchronization, the underlying symmetry group needs to have a real and orthogonal matrix representation. 
%
%We therefore convert $ISO(2)$, the group of two-dimensional translations and rotations, to $SO(3)$, the group of three-dimensional rotations. 
%
We then convert the two-dimensional translation and rotation to a three dimensional rotation. 
%
We first compute the Euler angles $\alpha_{ij}$, $\beta_{ij}$, and $\gamma_{ij}$,
\begin{equation} \label{eq:angle_relations}
%\begin{aligned}
%	\alpha_{ij} &= \theta_{ij} \\
%	\beta_{ij} &= \frac{dx_{ij}}{n} \times \eta_{proj} \\
%	\gamma_{ij} &= \frac{dy_{ij}}{n} \times \eta_{proj} \\
%\end{aligned}
	\alpha_{ij} = \theta_{ij}, \; \; \;
	\beta_{ij} = \frac{dx_{ij}}{n} \times \eta_{proj}, \; \; \;
	\gamma_{ij} = \frac{dy_{ij}}{n} \times \eta_{proj} 
\end{equation}
where $\eta_{proj}$ is the angular portion of the sphere onto which we choose to project the image.
%
We take $\eta_{proj} =  \pi/8$, so the image lies on a $\pi/8 \times \pi/8$ radians portion of the unit sphere.
%
This portion of the sphere is small enough so that curvature effects are not seen in the image.
%
We write rotations around the three principal axes, $R^x(\alpha)$, $R^y(\beta)$, and $R^z(\gamma)$, in terms of the three Euler angles
%\small
\begin{equation}
\begin{aligned}
	R^x(\alpha) &= \begin{bmatrix}
	1 & 0 & 0 \\
    0 & \cos(\alpha) & -\sin(\alpha) \\
    0 & \sin(\alpha) & \cos(\alpha)
	\end{bmatrix} \\
	R^y(\beta) &= \begin{bmatrix}
	\cos(\beta) & 0 & \sin(\beta) \\
    0 & 1 & 0 \\
    -\sin(\beta) & 0 & \cos(\beta)
    \end{bmatrix} \\
	R^z(\gamma) &= \begin{bmatrix} 
	\cos(\gamma) & -\sin(\gamma) & 0 \\
    \sin(\gamma) & \cos(\gamma) & 0 \\
    0 & 0 & 1 
    \end{bmatrix}.
\end{aligned}
\end{equation}
%\normalsize
%
The total rotation, $R_{ij} \in SO(3)$, is 
\begin{equation} \label{eq:total_rot}
	R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})	 = R^z(\gamma_{ij})  R^y(\beta_{ij})  R^x(\alpha_{ij})
%R(\alpha_{ij}, \beta_{ij}, \gamma_{ij}) = 
%\begin{bmatrix}
%\cos(\beta) \cos(\gamma) & \cos(\gamma)\sin(\alpha)\sin(\beta)-\cos(\alpha) \sin(\gamma) & \cos(\alpha)\cos(\gamma)\sin(\beta)+ \sin(\alpha)\sin(\gamma) \\
%\cos(\beta) \sin(\gamma) & \cos(\alpha) \cos(\gamma)+\sin(\alpha) \sin(\beta) \sin(\gamma) & \cos(\alpha) \sin(\beta) \sin(\gamma)-\cos(\gamma) \sin(\alpha) \\
%-\sin(\beta) & \cos(\beta) \sin(\alpha) & \cos(\alpha) \cos(\beta))
%\end{bmatrix}
\end{equation}
%
%\eqref{eq:total_rot} corresponds to first rotating the image by $\theta$, then translating the image by $dx$ pixels in the $x$ direction, and finally translating the image by $dy$ pixels in the $y$ direction (we note that, because the rotation matrices operate from the left, the rightmost rotation matrix in the product in \eqref{eq:total_rot} corresponds to the first operation we perform on the image).

From the rotation matrices $R(\alpha_{ij}, \beta_{ij}, \gamma_{ij})$,
we build the matrix $H$ in \eqref{eq:H_to_R} and compute the estimates $R_{i, est}$ in \eqref{eq:R_est}.
%
From the matrices $R_{1, est}, \dots, R_{m, est}$, we then find the corresponding translations and rotations of the images.
%
We first multiply all rotations by $R_{1, est}^T$ to ensure that all the images are (approximately) in the region of the sphere where we began.
%
We then compute the Euler angles $\alpha$, $\beta$, and $\gamma$ from the rotation matrix $R$ using the following relationships
\begin{equation}
\begin{aligned}
R_{1,1} & = \cos(\beta)\cos(\gamma) &
R_{2,1} & = \cos(\beta)\sin(\gamma) \\
R_{3,1} & = -\sin(\beta) &
R_{3,2} & = \sin(\alpha)\cos(\beta) \\
R_{3,3} & = \cos(\alpha)\cos(\beta)  &
\end{aligned}
\end{equation}
%
We then invert \eqref{eq:angle_relations} to compute the optimal translation and rotation for each image.
%
We round the optimal translations to the nearest integer to remove the need for interpolation when translating the images.
